---
title: Joins and data in the wild
teaching: 30
exercises: 15
questions:
- "How do I join data-sets?"
objectives:
- "To be able to join tibbles"
- "To understand the difference between the different join types"
- "To have an appreciation of some of the difficulties that can occur when using real-life data"
keypoints:
- "Use join_{left,inner,right} to join data on common values"
source: Rmd
---


```{r, include=FALSE}
source("../bin/chunk-options.R")
library(tidyverse)
knitr_fig_path("13-")
# Silently load in the data so the rest of the lesson works
gapminder <- read_csv("data/gapminder-FiveYearData.csv")
```


In this episode we will look at how to join tibbles together.  We will continue to use the gapminder data, and will look at how to join data on the medal performance of countries in the 2008 Olympics.  In doing so, we'll see some of the difficulties in working with real-world data, and will think about ways of solving them.

There are many ways of joining data in R; in common with the rest of this course we will focus on the tidyverse way, and use `dplyr` to join data.

## Loading the Olympic medal data


```{r, echo=FALSE, eval=FALSE}
# Olympic medal table - all the loading and cleaning will 
# be done outwith the lesson.  Just get participants to load 
# cleaned version.  This chunk will produce the required
# data file; it isn't run or visible in the lesson

library(rvest)
library(stringr)

medals <- read_html("https://en.wikipedia.org/wiki/2008_Summer_Olympics_medal_table") %>% 
  html_nodes(xpath='//*[@id="mw-content-text"]/div/table[2]') %>% html_table(fill=TRUE)
medals <- medals[[1]]

medals <- medals %>%
  filter(Rank != "Total (87 NOCs)") %>% 
  mutate(country=stringr::word(NOC,1, sep=fixed(intToUtf8(160)))) %>% # Delimited on &nbsp
  select(country, Gold, Silver, Bronze) %>% 
  rename(gold=Gold, silver=Silver, bronze=Bronze) 
write_csv(medals, "data/medals.csv")
```

The [medal table for the 2008 Olympic games was downloaded from Wikipedia](https://en.wikipedia.org/wiki/2008_Summer_Olympics_medal_table) and converted to a csv file.  We can load this in the same way as the gapminder data:

```{r}
medals <- read_csv("data/medals.csv")
```

We see that the column specification that `read_csv()` guessed looks reasonable; the countries are characters and columns containing the numbers of gold, silver and bronze medals have all been read as integers.

Before we join the two tables, complete the following challenges, which will generate a useful additional variable, and limit our gapminder data to the year closest to the medal data:

> ## Challenge: Generating medal totals
> 
> Use the `mutate` command to add a new column in your data-set
> called `total` that contains the total number of models awarded to 
> each country.  
>
> > ## Solution
> > 
> > ```{r}
> > medals <- medals %>% mutate(total = gold + silver + bronze)
> > ```
> > 
> {: .solution}
{: .challenge}

> ## Challenge: Filtering the gapminder data
> 
> Use `dplyr` to create a new tibble, gap2007,  that only contains data for 2007;
> the year in our gapminder data-set that is closest to 2008.
> 
> > ## Solution
> > 
> > ```{r}
> > gap2007 <- gapminder %>% filter(year == 2007)
> > ```
> > 
> {: .solution}
{: .challenge}

## Joining data tables

Before we join the `medals` and `gap2007` data, let's consider a smaller
example of the different ways we can join data.  The two data-sets we'll use are shown below:

```{r,echo = FALSE}

library(tibble)
animallegs <- tribble(~animal, ~legs,
                      "Cat", 4,
                      "Dog", 4,
                      "Bird", 2,
                      "Sheep", 4)

animalsound <- tribble(~animal, ~sound, 
                       "Cat", "miaw",
                       "Sheep", "baa",
                       "Dog", "growl",
                       "Bird", "tweet"
                        )
print(animallegs)
print(animalsound)
```

We see that the same animals appear in the "legs" and "sound" tables, albeit in a different order.
Let's join these using an _inner join_.  This will join on the variable we specify, if _both_ tables
have an observation with the same value:

```{r}
animalsoundlegs <- animallegs %>% inner_join(animalsound, by = "animal")
print(animalsoundlegs)
```

> ## Natural joins
>
> If we omit the `by` parameter from the join function, by default it will perform
> a _natural join_.  This means that the join will be performed on all variables with the 
> same name in both tables.  A message will be printed indicating the columns that
> have been used for the join.
>
> Do you think it is a good idea to make use of this feature?   Can you see any similarities
> to the discussion of the `col_types` argument of the `read_csv()` function?
> 
> > ## Solution
> > 
> > Although using natural joins can be a time-saver, it's safer to be explicit about 
> > the column(s) you're joining on.  It makes what your code is doing clearer to others
> > (and to you, when you revisit it).  If the format of your data changes, it will help
> > help prevent the code's behaviour changing. 
> > 
> > As with the `read_csv()` function, the join function prints out a message of the 
> > columns it has used to perform the join.  This is in the correct format to be copied and
> > pasted as a parameter to the function.  This can be a useful way of saving typing; you
> > can use natural joins while developing your code, check it's behaving as you expect,
> > and then convert them to named joins by pasting the `by = ...` message as a parameter 
> > to the join function.
> > 
> {: .solution}
{: .discussion}

## Left, right and full joins

The data tables we are joining  often won't contain the same observations.  To explore what happens when
we join data like this, let's load look at an extended version of the animal sounddata:

```{r, include = FALSE, message = FALSE}
# Write the example data-sets

animalsound <- animalsound %>% bind_rows(list(animal=c("Dog", "Lion"), sound=c("woof", "roar")))  %>% 
  filter(animal != "Sheep")
write_csv(animalsound, "data/animalsound.csv")
write_csv(animallegs, "data/animallegs.csv")
```

```{r}
animalsound <- read_csv("data/animalsound.csv",
                        col_types = cols(
                           animal = col_character(),
                           sound = col_character()
                        )
)

animallegs <- read_csv("data/animallegs.csv",
                       col_types =
                         cols(
                           animal = col_character(),
                           legs = col_integer()
                         )
)

print(animalsound)
print(animallegs)

```

In the new data, dogs make two noises ("growl" and "woof").  Although sheep is included in the legs table, there is no corresponding entry in the sound table.  Lion is included in the sound table, but does not appear in the legs table.

> ## Challenge
>
> Use `inner_join` to join the two tables.  Can you explain the entries in the resulting tibble?
>
> > ## Solution
> > We can join the data in the same way as in the example:
> > ```{r}
> > animallegs %>% inner_join(animalsound, by = "animal")
> > ```
> >
> > The observation for sheep has been dropped; an inner join returns a row if the data can 
> > if found in _both_ tables.  
> > 
> > There is a one to many relationship for the sounds a dog can make.  We obtain an observation 
> > for each of these. 
> > 
> {: .solution}
{: .challenge}

As we saw in the challenge, there is a row in our results table for _each_ of the sounds a dog can make.  The `join` functions will return the [cartesian product](https://en.wikipedia.org/wiki/Cartesian_product) of the matching rows.  In other words, if we are performing a 1 to many, or many to many match, a row for _each_ matching combination of variables will be returned.

The `inner_join()` function only returns rows if the matching variable appears in _both_ tables being joined.  An outer join will return rows if the data appears in _either_ table being joined.  Things get a little more complex here; does the data need to appear in _any_ of the tables, or does it matter _which_ of the tables the data is in?

An `full_join()` will return a row if the data appears in _any_ of the tables being joined:

```{r}
animalsound %>% full_join(animallegs, by = "animal")
```

As "Sheep" appears in the legs table, but not the sound table, its value of sound is `NA`. "Lion" appears in the sound table, but not in the legs table; its value  of legs is `NA`.

A `left_join()` will return a row if the data appears in the left hand table we're joining; this is the table that appears first in the command:

```{r}
animalsound %>% left_join(animallegs, by = "animal")
```

A row is returned for each of the animals in sound table, matched to the legs table. If there isn't a corresponding entry in the legs table, `NA`s are returned for columns coming from the legs table.

> ## Challenge
> 
> Use `right_join()` to join `animalsound` and `animallegs`.    Can you reproduce this result using only a `left_join()`?
> (don't worry about the order of the columns in the resulting tibble)
> 
> > ## Solution
> > ```{r}
> > animalsound %>% right_join(animallegs, by = "animal")
> > ```
> > 
> > Rows are returned for _each_ row in the legs table.  If there is no corresponding entry in the sound table,
> > `NA` is returned for these rows.
> >
> > If we reverse the order of the tables we can obtain the same result using a `left_join()`:
> >
> > ```{r}
> > animallegs %>% left_join(animalsound, by = "animal")
> > ```
> >
> > Note that the order of the columns in the data are different; the join functions will
> > put columns from the right hand table after the columns from the left.   It is good practice to refer
> > to columns by name, rather than position; this is one of the reasons why.
> > 
> {: .solution}
{: .challenge}

##  More complex joins

FIXME Write this section.  Need to cover joins on variables with different names, joins on multiple variables


## The medals data

```{r}

medaljoin <- gapminder %>% 
  filter(year == 2007) %>% 
  inner_join(medals, by="country") # Illustrate difference between join types
```
```{r}
medaljoin %>% ggplot(aes(x=gdpPercap, y=medalTotal, colour = continent)) + 
  geom_point()
```

